Beating Simple Backtracking Search in Subgraph Algorithms\thanks{This work was supported by the Engineering and Physical Sciences Research Council [grant numbers EP/026842/1 and EP/M508056/1].}

### 2018-11-27 paper/paper.tex
More words

A \emph{discrepancy} is where search goes against a value-ordering heuristic.  Discrepancy searches
\cite{DBLP:conf/ijcai/HarveyG95,DBLP:conf/aaai/Korf96,DBLP:conf/ijcai/Walsh97,DBLP:conf/cpaior/KarouiHLN07}
are alternatives to backtracking search that initially search disallowing all discrepancies, and
then retry search allowing an increasing number of discrepancies at each iteration until either a
solution is found or unsatisfiability is proven. These schemes assume that value-ordering heuristics
are usually reliable, and that most solutions can be found with only a small number of
discrepancies; in such cases, the heavy commitment to early branching choices made by conventional
backtracking search can be extremely costly.

@@

But what happens if our value-ordering heuristic has to choose between mapping a pattern vertex to,
for example, three target vertices of degree ten, two vertices of degree nine, or five of degree
two? When driving conventional backtracking search, the degree heuristic would pick one of the
vertices of degree ten, and we would commit all of our search effort to the exponentially large
search tree underneath it, not considering any other choice until this tree has been fully explored
and eliminated.  We will show that this is not a wise choice, and that instead, we should
\emph{commit equal search effort} to each of the three subproblems found by mapping to vertices of
degree ten.  And similarly, should we be certain that vertex of degree ten is so much better than a
vertex of degree nine that we should commit no effort to degree nine vertices until all the degree
ten subproblems have been explored? Or might it be better to commit, say, twice as much effort to
each degree ten subproblems as to each degree nine subproblems, and only a very small amount of
effort to the degree two subproblems?  To test this hypothesis, we will now introduce a new
alternative to backtracking search, which we call \emph{solution-biased search}.  This search is
made up of three components: a new slightly-random value-ordering heuristic, rapid restarts, and
nogood recording. The aim is to perform a complete search, but spending proportionally more time in
parts of the search tree that are preferred by the value-ordering heuristic.

@@

\Cref{figure:value-ordering-heuristics} confirms that this heuristic, when used with backtracking
search, will solve close to the same number of instances as the degree heuristic would for any given
choice of timeout. In other words, we can introduce an element of randomness into the degree
value-ordering heuristic without adversely affecting its performance \emph{in aggregate}.  The
right-hand plot gives a detailed comparison. It shows that despite the aggregate performance being
similar, on a case by case basis, the two heuristics can make a large difference to the performance
for individual satisfiable instances. This justifies our belief that although degree is a good
heuristic, we should perhaps not commit heavily to a vertex of highest degree, but also consider
vertices of the same or similar degree.

@@

\subsection{Why ``Solution-Biased'', and why Exponential Bias?}

?? Although we have shown that it provides good results, we have yet to justify where the biased
formula comes from, or indeed why we call this approach ``solution-biased''. The crux of the
matter is that whilst we do not know where solutions actually lie in the search space, we \emph{do}
know that subproblems under a target vertex of degree $d$ tend to contain a small constant times
more search nodes than those under a target vertex of degree $d - 1$.

@@

We have shown that it is both possible and beneficial to introduce a small amount of randomness into
backtracking search algorithms for hard subgraph problems, without compromising existing search
order heuristics, and without paying a penalty for unsatisfiable instances. In principle it is
well-known that introducing some form of randomness \cite{DBLP:conf/aaai/GomesSK98} or reduced
commitment \cite{DBLP:conf/ijcai/HarveyG95} into backtracking search can be beneficial, but until
now it has not been clear how to do so efficiently in practice, particularly when we already have
strong search ordering heuristics and when are expecting to encounter large numbers of unsatisfiable
instances. Our results show that the key to success is a combination of aggressive restarts, nogood
recording, and biasing an existing value-ordering heuristic with just a pinch of randomness.

>>>>

A \emph{discrepancy} is where search goes against the advice of a value-ordering heuristic.
Discrepancy searches
\cite{DBLP:conf/ijcai/HarveyG95,DBLP:conf/aaai/Korf96,DBLP:conf/ijcai/Walsh97,DBLP:conf/cpaior/KarouiHLN07}
are alternatives to backtracking search that initially search disallowing all discrepancies, and
then retry search allowing an increasing number of discrepancies at each iteration until either a
solution is found or unsatisfiability is proven. These schemes assume that value-ordering heuristics
are usually reliable, and that most solutions can be found with only a small number of
discrepancies; in such cases, the heavy commitment to early branching choices made by conventional
backtracking search can be extremely costly.

@@

But what happens if our value-ordering heuristic has to choose between mapping a pattern vertex to
one of, for example, three target vertices of degree ten, two vertices of degree nine, or five of
degree two? When driving conventional backtracking search, the degree heuristic would pick one of
the vertices of degree ten, and we would commit all of our search effort to the exponentially large
search tree underneath it, not considering any other choice until this tree has been fully explored
and eliminated.  We will show that this is not a wise choice, and that instead, we should
\emph{commit equal search effort} to each of the three subproblems found by mapping to vertices of
degree ten.  And similarly, should we be certain that vertex of degree ten is so much better than a
vertex of degree nine that we should commit no effort to degree nine vertices until all the degree
ten subproblems have been explored? Or might it be better to commit, say, twice as much effort to
each degree ten subproblems as to each degree nine subproblems, and only a very small amount of
effort to the degree two subproblems?  To test this hypothesis, we will now introduce a new
alternative to backtracking search, which we call \emph{solution-biased search}.  This search is
made up of three components: a new slightly-random value-ordering heuristic, rapid restarts, and
nogood recording. The aim is to perform a complete search, but spending proportionally more time in
parts of the search tree that are preferred by the value-ordering heuristic.

@@

\Cref{figure:value-ordering-heuristics} confirms that this heuristic, when used with backtracking
search, will solve close to the same number of instances as the degree heuristic would for any given
choice of timeout. In other words, we can introduce an element of randomness into the degree
value-ordering heuristic without adversely affecting its performance \emph{in aggregate}.  The
right-hand plot gives a detailed comparison. It shows that despite the aggregate performance being
similar, on a case by case basis, the two heuristics can make a large difference to the performance
for individual satisfiable instances. This justifies our belief that although degree is a good
heuristic, we should perhaps not commit heavily to a single vertex of highest degree, but also consider
vertices of the same or similar degree.

@@

\subsection{Solution-Biased Search in Theory}

Although we have shown that it provides good results, we have yet to justify where the biased
formula comes from, or indeed why we call this approach ``solution-biased''.  Our goal is to use
biased randomness in a value-ordering heuristic to spend time in subproblems proportional to an
estimate of their solution density \cite{DBLP:journals/jair/PesantQZ12}. ?? I feel like we still
need a sentence here saying why this is a good idea.

To estimate solution density, we need an estimate of how big different subproblems are likely to be,
and of how many solutions each subproblem is likely to contain. Of course, obtaining exact (or even
approximate) values for these figures is at least as hard as solving the problem in its entirety.
For pairs of Erd\H{o}s-R\'enyi
random graphs with large solution counts (i.e.\ chosen from within the ``easy satisfiable'' region
\cite{DBLP:journals/jair/McCreeshPST18}), we observe a linear relationship between subproblem size
and number of solutions. Thus, for graphs from this distribution, we need only an estimator of
subproblem size. Measurements also suggest that, for pairs of Erd\H{o}s-R\'enyi graphs, subproblems
under a target vertex of degree $d$ tend to contain a small constant times more search nodes than
those under a target vertex of degree $d - 1$.  This empirical analysis suggests that an estimator
that is exponential in $d$ will give our method the desired behaviour, at least for
Erd\H{o}s-R\'enyi graphs. We expect it may be possible to derive better estimators for particular
input classes, although over the full range of problem instances, we have verified that estimators
of the form $e^d$ substantially outperform polynomial and factorial weightings.

@@

We have shown that it is both possible and beneficial to introduce a small amount of randomness into
backtracking search algorithms for hard subgraph problems, without compromising existing search
order heuristics, and without paying a penalty for unsatisfiable instances. In principle it is
well-known that introducing some form of randomness \cite{DBLP:conf/aaai/GomesSK98} or reduced
commitment \cite{DBLP:conf/ijcai/HarveyG95} into backtracking search can be beneficial, but until
now it has not been clear how to do so efficiently in practice, particularly when we already have
strong search ordering heuristics and when we are expecting to encounter large numbers of unsatisfiable
instances. Our results show that the key to success is a combination of aggressive restarts, nogood
recording, and biasing an existing value-ordering heuristic with just a pinch of randomness.
